\documentclass[12pt]{article}

\include{qun-sheet-texhead}
\include{diagrams}

\title{Sheet 3: Quaternions and rotations}
\author{J. Evans}
\date{}

\begin{document}
\maketitle

\bigskip

{\bf The final mark out of 10 (which counts towards your grade) will be calculated as your mark on Q1 plus your mark on Q2 plus your best mark from Q3--5.} I will also award stars: silver for a total mark of 12 or more on any four questions, gold for a total mark of 15 or more on all questions. 

\vspace{1cm}

\begin{question}\ (2 marks)\\
  In each case, find unit quaternions $q$ such that the map $\OP{Im}(\HH)\to\OP{Im}(\HH)$, $x\mapsto qxq^{-1}$, is the specified orthogonal transformation:
  \begin{enumerate}
  \item[(a)] $\left(\begin{array}{ccc}0 & -1 & 0\\ 1 & 0 & 0\\ 0 & 0 & 1\end{array}\right)$.
  \item[(b)] $\left(\begin{array}{ccc}1/\sqrt{2} & 0 & -1/\sqrt{2}\\ 0 & 1 & 0\\ 1/\sqrt{2} & 0 & 1/\sqrt{2}\end{array}\right)$.
  \end{enumerate}
\end{question}

\begin{answer}
  \begin{enumerate}
  \item[(a)] This first matrix is a $\pi/2$-rotation around the $z$-axis, so should be effected by the quaternion $\cos(\pi/4)+k\sin(\pi/4)=\frac{1}{\sqrt{2}}(1+k)$. We can check that this is indeed the case:
    \begin{align*}
      \frac{1}{\sqrt{2}}(1+k)(ix+jy+kz)\frac{1}{\sqrt{2}}(1-k)&=\frac{1}{2}((i+j)x+(j-i)y+(k-1)z)(1-k)\\
      &=\frac{1}{2}((i+j+j-i)x+(j-i-i-j)y+(k-1+1+k)z)\\
      &=-iy+jx+kz
    \end{align*}
    so $(1,0,0)$ is sent to $(0,1,0)$, $(0,1,0)$ is sent to $(-1,0,0)$ and $(0,0,1)$ is sent to $(0,0,1)$, as required.
  \item[(b)] This matrix corresponds to a $-\pi/4$-rotation around the $y$-axis, so should be effected by the quaternion $\cos(\pi/8)-j\sin(\pi/8)$.
  \end{enumerate}
\end{answer}
\newpage

\vspace{1cm}

\begin{question}\ (5 marks)\\
  \begin{enumerate}
  \item[(a)]  Let $q_1=t_1+ix_1+jy_1+kz_1$ and $q_2=t_2+ix_2+jy_2+kz_2$. Show that
  \[\OP{Re}(\bar{q}_1q_2)=t_1t_2+x_1x_2+y_1y_2+z_1z_2.\]
  \item[(b)] Suppose that $t_1=t_2=0$. Show that $q_1q_2=-q_1\cdot q_2+q_1\times q_2$, i.e. the real part equals $-(x_1x_2+y_1y_2+z_1z_2)$ and the imaginary part equals $i(y_1z_2-y_2z_1)+j(z_1x_2-z_2x_2)+k(x_1y_2-x_2y_1)$.
  \item[(c)] Show that $A\colon G\times G\to\OP{Maps}(\HH,\HH)$, $A(g,h)x=gxh^{-1}$ defines an action of $G\times G$ on $\HH$.
  \item[(d)] Show that $A(g)\colon x\mapsto gxh^{-1}$ is an isometry of $\HH$ when $g,h\in G$.
  \item[(e)] Parts (c) and (d) imply that we have a homomorphism $G\times G\to O(4)$. Show that the kernel of this homomorphism is $\{(1,1),(-1,-1)\}$. [Hint: Consider the effect of the isometry $x\mapsto gxh^{-1}$ on the quaternions $x=1$, $x=i$, $x=j$]
  \end{enumerate}
\end{question}
\begin{answer}
  \begin{enumerate}
  \item[(a)] We have
    \[(t_1-ix_1-jy_1-kz_1)(t_2+ix_2+jy_2+k_2)=t_1t_2-i^2x_1x_2-j^2y_1y_2-k^2z_1z_2+i(\cdots)+j(\cdots)+k(\cdots).\]
  \item[(b)] We have
    \[(ix_1+jy_1+kz_1)(ix_2+jy_2+kz_2)=-(x_1x_2+y_1y_2+z_1z_2)+ijx_1y_2+jiy_1x_2+\cdots\]
    and $ij=-ji=k$, etc.
  \item[(c)] We need to show that $A(1,1)(x)=1x1=x$ and that
    \begin{align*}
      A(g_1,h_1)A(g_2,h_2)x&=A(g_1,h_1)g_2xh_2^{-1})\\
      &=g_1g_2xh_2^{-1}h_1^{-1}\\
      &=g_1g_2x(h_1h_2)^{-1}\\
      &=A(g_1g_2,h_1h_2)x.
    \end{align*}
  \item[(d)] We have $|gxh^{-1}|=|g||x||h^{-1}|=|x|$ since $|g|=|h|=|h^{-1}|=1$.
  \item[(e)] $(g,h)$ is in the kernel of this homomorphism if $gxh^{-1}=x$ for all $x$, i.e. $gx=xh$. For $x=1$ this implies $g=h$. For $x=i$ this implies $gi=ig$. If $g=a+ib+jc+kd$ then $gi-ig=ai-b-kc+jd-ia+b-kc+jd=2(jd-kc)$ so $gi-ig=0$ implies $d=c=0$. Similarly, $gj=jg$ implies $b=0$. Therefore $g$ is real and has length 1, so it must be $\pm 1$. This leaves the two possibilities $(1,1)$ and $(-1,-1)$.
  \end{enumerate}
\end{answer}

\newpage

\begin{question}\ (3 marks)\\
  Suppose that $y$ is a unit quaternion and let $H\subset\HH$ be the plane orthogonal to $y$. Show that the reflection in the plane $H$ is given in terms of quaternionic multiplication by $z\mapsto -y\bar{z}y$.
\end{question}

\begin{answer}
  We know that $z\cdot y=\OP{Re}(\bar{z}y)$ so
  \begin{align*}
    z-2(z\cdot y)y&=z-2\OP{Re}(\bar{z}y)y\\
    &=z-y(\bar{z}y+\bar{y}z)y\\
    &=z-y\bar{z}y+y\bar{y}z\mbox{ as }y\mbox{ commutes with real numbers}\\
    &=-y\bar{z}y\mbox{ as }|y|=1.
  \end{align*}
\end{answer}
\newpage

\vspace{1cm}

\begin{question}\ (3 marks)
  Consider the 2-by-2 complex matrices:
  \[\sigma_1=\left(\begin{array}{cc}
    i & 0\\
    0 & -i
  \end{array}\right),\quad\sigma_2=\left(\begin{array}{cc}
    0 & 1\\
    -1 & 0
  \end{array}\right),\quad\sigma_3=\left(\begin{array}{cc}
    0 & i\\
    i & 0
  \end{array}\right).\]
  Show that $\sigma_1^2=\sigma_2^2=\sigma_3^2=-1$ and that $\sigma_a\sigma_b=\sigma_c=-\sigma_b\sigma_a$ where $a,b,c$ is a cyclic permutation of $1,2,3$.

  Consider the group of 2-by-2 complex matrices
  \[SU(2)=\{A\ :\ A^{\dagger}=A^{-1},\ \det A=1\}\mbox{ where }\dagger\mbox{ denotes conjugate transpose.}\]
  Show that if $\left(\begin{array}{cc}a & b\\ c & d\end{array}\right)\in SU(2)$ then $c=-\bar{b}$, $d=\bar{a}$ and $|a|^2+|b|^2=1$. If $a=t+ix$ and $b=y+iz$ then show that
    \[\left(\begin{array}{cc}a&b\\-\bar{b}&\bar{a}\end{array}\right)=t\mathbf{1}+x\sigma_1+y\sigma_2+z\sigma_3\]
    and deduce that $SU(2)$ is isomorphic to the group of unit quaternions.
\end{question}

\begin{answer}
  The first part is just a matrix computation. For the second part, observe that
  \[A^{\dagger}=\left(\begin{array}{cc}\bar{a} & \bar{c}\\\bar{b} & \bar{d}\end{array}\right),\qquad A^{-1}=\left(\begin{array}{cc}d & -b\\-c & a\end{array}\right)\]
    so if $A\in SU(2)$ then $c=-\bar{b}$ and $d=\bar{a}$. Moreover, $\det A=ad-bc=a\bar{a}+b\bar{b}=|a|^2+|b|^2=1$. The next line is just obtained by substituting the matrices $\sigma_m$ in and expanding. Finally, we use the isomorphism
    \[G\to SU(2),\qquad t+ix+jy+kz\mapsto t\mathbf{1}+x\sigma_1+y\sigma_2+z\sigma_3.\]
    This is clearly a bijection between the unit quaternions and the matrices in $SU(2)$ (the unit condition becomes the determinant one condition) so we just need to show it is a homomorphism. The multiplication
    \[\left(t_1+x_1\sigma_1+y_1\sigma_2+z_1\sigma_3\right)\left(t_2+x_2\sigma_1+y_2\sigma_2+z_2\sigma_3\right)\]
    is clearly bilinear in the coefficients and therefore it suffices to check that it gives the correct answer on a basis. This is what we did in the first part of the question.
\end{answer}
\newpage

\vspace{1cm}

\begin{question}\ (3 marks)\\
  We define the {\em exponential} of a purely imaginary quaternion $q$ by the formula
  \[e^q:=\sum_{m=0}^\infty\frac{1}{m!}q^m.\]
  If $q=u\theta$ for $|u|=1$, prove that $e^q=\cos\theta+u\sin\theta$.

  Check that
  \[e^{i\pi/2}e^{j\pi/2}\neq e^{(i+j)\pi/2}\]
  (i.e. the usual law of logarithms fails).
  
  [This is just the beginning of a much more general story of exponential maps for groups. For more, see the 4th year course {\bf Lie groups and Lie algebras}.]
\end{question}

\begin{answer}
  We have
  \begin{align*}
    e^{u\theta}&=1+u\theta+\frac{1}{2!}u^2\theta^2+\frac{1}{3!}u^3\theta^3+\cdots\\
    &=\left(1-\frac{1}{2!}\theta^2+\cdots\right)+u\left(\theta-\frac{1}{3!}\theta^3+\cdots\right)\\
    &=\cos\theta+u\sin\theta.
  \end{align*}
  since $u^2=-1$.

  For the second part we have $e^{i\pi/2}=i$, $e^{j\pi/2}=j$, $ij=k$ and $e^{(i+j)\pi/2}=e^{\frac{i+j}{\sqrt{2}}\frac{\pi}{\sqrt{2}}}=\cos\left(\frac{\pi}{\sqrt{2}}\right)+\frac{i+j}{\sqrt{2}}\sin\left(\frac{\pi}{\sqrt{2}}\right)$ so $e^{i\pi/2}e^{j\pi/2}\neq e^{(i+j)\pi/2}$.
\end{answer}
\newpage


\end{document}
